---
output: pdf_document
geometry: margin=1.0in
fontsize: 12pt
header-includes:
  - \usepackage{setspace}
  - \singlespacing 
  - \usepackage{nopageno} 
---

```{r}
library(tidyverse)
library(gssr)
```


```{r}
gss_get_marginals(varnames = "degree", data = gss_doc, margin = marginals)
```


```{r}
gss_2018 <- gss_all |> 
  filter(year == 2018) |> 
  select(id, mntlhlth, degree) |> 
  drop_na()
```

```{r}
gss_doc %>% filter(id == "degree") %>%
  select(marginals) %>% 
  unnest(cols = c(marginals))
```



### Questions

1. Does the mean number of days of poor mental health differ from 3 at the 99% confidence level?

```{r}
t.test(gss_2018$mntlhlth, mu = 3, conf.level = .99)
```


2. Among respondents reporting any days of poor mental health, does the mean number of days differ from 8.5 at the 95% confidence level?

```{r}
t.test(gss_2018$mntlhlth[gss_2018$mntlhlth>0], 
       mu = 8.5, conf.level = .95)
```

3. Consider respondents who have not entered or completed college. Among these respondents, does the proportion with any mental health days differ from .40 at the 99% confidence level?

```{r}
gss_2018 <- gss_2018 |> 
  mutate(mntlhlth_any = ifelse(mntlhlth>0, 1, 0))
```

```{r}
addmargins(table(gss_2018$degree, gss_2018$mntlhlth_any))
```

```{r}
prop.test(45+295, 120+671, p = .4, conf.level = .99)
```

4. Consider respondents who have completed college or more. Among these respondents, does the proportion with any mental health days differ from .40 at the 95% confidence level?


```{r}
addmargins(table(gss_2018$degree, gss_2018$mntlhlth_any))
```

```{r}
prop.test(151+70, 323+168, p = .4, conf.level = .95)
```

```{r}
t.test(gss_2018$mntlhlth[gss_2018$degree<=1 & gss_2018$mntlhlth>0], gss_2018$mntlhlth[gss_2018$degree>=3 & gss_2018$mntlhlth>0])
```

