## Setting Up

We'll start with the `gss_week8.csv` file again today. Load it with the usual packages.

```{r}
#| message: false

library(tidyverse)
gss_week8 <- read.csv("https://raw.githubusercontent.com/mjclawrence/soci385_f23/main/data/gss_week8.csv")
```


## Warm Up

1. Test if there is a significant difference (at the 95% confidence level) in average number of political actions between respondents who self identify as lower class and respondents who self identify as working class. Use the `polactions` variable for political actions and the `class` variable for subjective class identification. Start with a table of `class` to see the value labels.

### REPLACE THIS. LINE WITH YOURCODE CHUNK ###

2. Test if there is a significant difference (at the 95% confidence level) in average number of political actions (`polactions`) by sex (`sex`). Get a table of `sex` to see how the values are labeled.

### REPLACE THIS LINE WITH YOUR CODE CHUNK ###

3. Test if there is a significant difference (at the 95% confidence level) in average age between respondents who work for someone else and respondents who are self employed. Use the `age` variable for age and the `wrkslf` variable for work status. Use a table with the `wrkslf` variable to see how the values are labeled.

### REPLACE THIS LINE WITH YOUR CODE CHUNK ###

4. Test if there is a significant difference in the average age at first birth (`agekdbrn`) by two values of race (`racehisp`). Only use female respondents' age at first birth. Check tables of `sex` and `racehisp` to see how the values are labeled.

### REPLACE THIS LINE WITH YOUR CODE CHUNK ###



## Comparing Proportions

We have seen over the last couple weeks that proportions and means have different standard errors. That is still true when we are comparing differences in proportions. Also, when working with differences in proportions, we will use different SEs for confidence intervals and for significance tests.

The SE for a CI of a difference in proportions is very similar to what we used last week as the SE for a CI of a proportion:

$\Large{se = \sqrt{ \frac {\hat{\pi}_1 (1 - \hat{\pi}_1)} {n_1} + \frac {\hat{\pi}_2 (1 - \hat{\pi}_2)} {n_2} }}$

The SE for difference uses a pooled estimate of the proportion and the n's for each sample:

$\Large{se = \sqrt{ \hat{\pi}(1-\hat{\pi})  (\frac {1} {n_1} + \frac{1}{n_2})  }}$

The null hypothesis is that proportions are the same in each sample. Formally, this is written as:

$\Large{\pi_2 - \pi_1 = 0}$

The test statistic is calculated as:

$\Large{\frac{\hat{\pi}_1 - {\hat{\pi}_2}}{se}}$


## Example

We'll use data from an audit study examining discrimination by race and sex in the labor market. The researchers - Marianne Bertrand and Sendhil Mullainathan - described their study this way: "We [responded] to over 1300 employment ads in the sales, administrative support, clerical and customer services job categories and [sent] nearly 5000 resumes. The ads we [responded] to cover a large spectrum of job quality, from cashier work at retail establishments and clerical work in a mailroom to office and sales management positions." 

They responded to ads by submitting fictitious resumes with similar educational backgrounds and job experiences, but they varied the race (black/white) and sex (female/male) of the applicants' names. They then recorded which resumes received call backs. What is your hypothesis for which race-sex combination received the highest and lowest proportions of call backs?

Let's load the data:

```{r}
resume <- read_csv("https://raw.githubusercontent.com/mjclawrence/soci385_f23/main/data/resume.csv")
```

Open the spreadsheet view of the data to see how the names vary by race and sex.

What proportions of resumes with white and black names received call backs?

### REPLACE THIS LINE WITH YOUR CODE CHUNK ###

We want to test if the *difference* between these two proportions is significant. When we wanted to test if one proportion differed from a null hypothesis value, we fed prop.test() the number in our sample with a 1 for a binary variable and the total number in our sample. To test differences in proportions, we will feed prop.test() a cross table (or contingency table).

The table we need will have the `race` and `call` variables. Save the table as an object so you can use it later. Note that we do not need to find the sums of each row with addmargins().

```{r}
race_table <- table(resume$race, resume$call)
```

You can see the frequencies by recalling the object name.

```{r}
race_table
```

You don't have to redo the table to get proportions. Simply wrap the object name in the `prop.table()` function (with a ",1" for row proportions).

```{r}
prop.table(race_table, 1)
```

The proportions in the 1 column are the same that we calculated earlier showing the proportion of resumes with black and white names that received call backs. Feed the saved table - the one with frequencies, not proportions - into prop.test() to test if the difference in these proportions is significant. More precisely, we are testing if the difference in proportions significantly differs from a null hypothesis value of zero (which would tell us there is no difference in the proportions).

```{r}
prop.test(race_table)
```

The sample estimates for "prop 1" and "prop 2" refer to the proportions with 0 in our table. Subtracting those proportions from 1 would give you the proportions with 1 in our table. Prop 1 is for the first sample you listed in your t.test() function and Prop 2 is for the second sample in your t.test() function.

Can we reject the null hypothesis? Is the difference significant?

Here is an example from the paper[^1] of how to write up the results:

> "We find large racial differences in callback rates. Applicants with White names need to send about 10 resumes to get one callback whereas applicants with African American names need to send around 15 resumes to get one callback. This 50 percent gap in callback rates in statistically very significant. Based on our estimates, a White name yields as many more callbacks as an additional eight years of experience. Since applicants' names are randomly assigned, this gap can only be attributed to the name manipulation."

[^1]: Bertrand, Marianne and Sendhil Mullainathan. 2004. "Are Emily and Greg More Employable Than Lakisha and Jamal? A Field Experiment on Labor Market Discrimination." *American Economic Review* 94(4): 991-1013.

Now test if there is a significant difference in the proportion of call backs between resumes with female and male names.

### REPLACE THIS LINE WITH YOUR CODE CHUNK ###


How would we get the proportions of call backs by race *and* sex?

### REPLACE THIS LINE WITH YOUR CODE CHUNK ###


In groups, let's look at four more differences:

- Differences in the proportions of call backs for resumes with black female names vs resumes with white female names
- Differences in the proportions of call backs for resumes with white male names vs resumes with black male names
- Differences in the proportions of call backs for resumes with white female names vs resumes with white male names
- Differences in the proportions of call backs for resumes with black female names vs resumes with black male names

### REPLACE THIS LINE WITH YOUR CODE CHUNK ###


## Bonus! 

What names had the highest and lowest chances of receiving call backs?

### REPLACE THIS LINE WITH YOUR CODE CHUNK ###

