---
title: "Differences in Proportions"
author: "Matt Lawrence"
date: "11/2/2023"
format: html
---

## Setting Up

We'll start with the `gss_week8.csv` file again today. Load it with the usual packages.

```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
gss_week8 <- read.csv("https://raw.githubusercontent.com/mjclawrence/soci385_f23/main/data/gss_week8.csv")
```


## Warm Up

1. Test if there is a significant difference (at the 95% confidence level) in average number of political actions between respondents who self identify as lower class and respondents who self identify as working class. Use the `polactions` variable for political actions and the `class` variable for subjective class identification. Start with a table of `class` to see the value labels.

### REPLACE THIS. LINE WITH YOURCODE CHUNK ###

```{r}
table(gss_week8$class)
```

```{r}
# Average number of political actions between lower and working class respondents
t.test(gss_week8$polactions[gss_week8$class=="Lower"],
       gss_week8$polactions[gss_week8$class=="Working"],
       conf.level = .95)
```

Difference is not significant because you cannot reject the null hypothesis that the true difference is 0. Test statistic of -1.5848 is not more extreme than -1.96, p-value of .1142 is not less than .05, and confidence interval does include 0.

What is the relationship between the test statistic and the p-value?

```{r}
pt(-1.5848, df = 268.46) # probability of a more extreme value to the left of negative t
1 - pt(1.5848, df = 268.46) # probability of a more extreme value to the right of positive t
pt(-1.5848, df = 268.46) + (1 - pt(1.5848, df = 268.46)) # sum of probabilities = p-value
pt(-1.5848, df = 268.46) * 2 # simplified way to get the sum of probabilities
```

These degrees of freedom are different from what we saw before because the t-test adjusts for the samples' variances not being equal. If we assert that the variances are equal, we get df = n1 + n2 - 2

```{r}
t.test(gss_week8$polactions[gss_week8$class=="Lower"],
       gss_week8$polactions[gss_week8$class=="Working"],
       conf.level = .95,
       var.equal = TRUE)
```

2. Test if there is a significant difference (at the 95% confidence level) in average number of political actions (`polactions`) by sex (`sex`). Get a table of `sex` to see how the values are labeled.

### REPLACE THIS LINE WITH YOUR CODE CHUNK ###

```{r}
t.test(gss_week8$polactions[gss_week8$sex == "Female"],
       gss_week8$polactions[gss_week8$sex == "Male"])
```

Cannot reject the null hypothesis; difference is not significant. Test statistic is not more extreme than -1.96, p-value is not less than .05, and the null hypothesis value of 0 is within the confidence interval.

3. Test if there is a significant difference (at the 95% confidence level) in average age between respondents who work for someone else and respondents who are self employed. Use the `age` variable for age and the `wrkslf` variable for work status. Use a table with the `wrkslf` variable to see how the values are labeled.

### REPLACE THIS LINE WITH YOUR CODE CHUNK ###

```{r}
# Average age between respondents who work for someone else and who are self employed
t.test(gss_week8$age[gss_week8$wrkslf=="Work For Someone Else"],
       gss_week8$age[gss_week8$wrkslf=="Self Employed"])
```

Difference is significant because you can reject the null hypothesis that the true difference is 0. Test statistic of -12.416 is more extreme than -1.96, p-value of 2.2e-16 is less than .05, and confidence interval does not include 0.

4. Age at first birth (`agekdbrn`) by two values of race (`racehisp`). Only use female respondents' age at first birth. Check a table of `sex` and `racehisp` to see how the values are labeled.

### REPLACE THIS LINE WITH YOUR CODE CHUNK ###

```{r}
t.test(gss_week8$agekdbrn[gss_week8$racehisp=="Black" & gss_week8$sex=="Female"],
       gss_week8$agekdbrn[gss_week8$racehisp=="Hispanic" & gss_week8$sex=="Female"])
```


## Comparing Proportions

We have seen over the last couple weeks that proportions and means have different standard errors. That is still true when we are comparing differences in proportions. Also, when working with differences in proportions, we will use different SEs for confidence intervals and for significance tests.

The SE for a CI of a difference in proportions is very similar to what we used last week as the SE for a CI of a proportion:

$\Large{se = \sqrt{ \frac {\hat{\pi}_1 (1 - \hat{\pi}_1)} {n_1} + \frac {\hat{\pi}_2 (1 - \hat{\pi}_2)} {n_2} }}$

The SE for difference uses a pooled estimate of the proportion and the n's for each sample:

$\Large{se = \sqrt{ \hat{\pi}(1-\hat{\pi})  (\frac {1} {n_1} + \frac{1}{n_2})  }}$

The null hypothesis is that proportions are the same in each sample. Formally, this is written as:

$\Large{\pi_2 - \pi_1 = 0}$

The test statistic is calculated as:

$\Large{\frac{\hat{\pi}_1 - {\hat{\pi}_2}}{se}}$


## Example

We'll use data from an audit study examining discrimination by race and sex in the labor market. Researchers responded to job ads for positions in sales, administrative support, clerical fields, and customer service. They submitted fictitious resumes with similar educational backgrounds and job experiences, but they varied the race (black/white) and sex (female/male) of the names. They then recorded which resumes received call backs. What is your hypothesis for which race-sex combination received the highest and lowest proportions of call backs?

Let's load the data:

```{r}
resume <- read_csv("https://raw.githubusercontent.com/mjclawrence/soci385_f23/main/data/resume.csv")
```

Open the spreadsheet view of the data to see how the names vary by race and sex.

What proportion of resumes with white and black names received call backs?

### REPLACE THIS LINE WITH YOUR CODE CHUNK ###

```{r}
resume |> 
  group_by(race) |> 
  summarise(prop_callback = mean(call))
```

We want to test if the *difference* between these two proportions is significant. When we wanted to test if one proportion differed from a null hypothesis value, we fed prop.test() the number in our sample with a 1 for a binary variable and the total number in our sample. To test differences in proportions, we will feed prop.test() a cross table (or contingency table).

The table we need will have the `race` and `call` variables. Save the table as an object so you can use it later. Note that we do not need to find the sums of each row with addmargins.

```{r}
race_table <- table(resume$race, resume$call)
```

You can see the frequencies by recalling the object name.

```{r}
race_table
```

You don't have to redo the table to get proportions. Simply wrap the object name in the `prop.table()` function (with a ",1" for row proportions).

```{r}
prop.table(race_table, 1)
```

The proportions in the 1 column are the same that we calculated earlier showing the proportion of resumes with black and white names that received call backs. Feed the saved table - the one with frequencies, not proportions - into prop.test() to test if the difference in these proportions is significant. More precisely, we are testing if the difference in proportions significantly differs from a null hypothesis value of zero (which would tell us there is no difference in the proportions).

```{r}
prop.test(race_table)
```

The sample estimates for "prop 1" and "prop 2" refer to the proportions with 0 in our table. Subtracting those proportions from 1 would give you the proportions with 1 in our table.

Can we reject the null hypothesis? Is the difference significant?

Now test if there is a significant difference in the proportion of call backs between resumes with female and male names.

### REPLACE THIS LINE WITH YOUR CODE CHUNK ###

```{r}
sex_table <- table(resume$sex, resume$call)
prop.table(sex_table,1)
```

```{r}
prop.test(sex_table)
```

We cannot reject the null hypothesis. The p-value is not less than .05 and the 95 percent confidence interval includes the null hypothesis value of 0.

How would we get the proportions by race *and* sex?

### REPLACE THIS LINE WITH YOUR CODE CHUNK ###

```{r}
resume |> 
  group_by(race, sex) |> 
  summarise(prop_callback = mean(call)) |> 
  arrange(-prop_callback)
```

In groups, let's look at four more differences:

- Differences in the proportions of call backs for resumes with black female names vs resumes with white female names
- Differences in the proportions of call backs for resumes with white male names vs resumes with black male names
- Differences in the proportions of call backs for resumes with white female names vs resumes with white male names
- Differences in the proportions of call backs for resumes with black female names vs resumes with black male names



### Race for females is significant

```{r}
race_female_table <- table(resume$race[resume$sex=="female"],
                           resume$call[resume$sex=="female"])
prop.test(race_female_table)
```

### Race for males is not significant

```{r}
race_male_table <- table(resume$race[resume$sex=="male"],
                        resume$call[resume$sex=="male"])
prop.test(race_male_table)
```

### Sex for whites is not significant

```{r}
sex_white_table <- table(resume$sex[resume$race=="white"],
                         resume$call[resume$race=="white"])
prop.test(sex_white_table)
```


### Sex for blacks is not significant

```{r}
sex_black_table <- table(resume$sex[resume$race=="black"],
                         resume$call[resume$race=="black"])
prop.test(sex_black_table)
```


Bonus! What names had the highest and lowest chances of receiving call backs?

### REPLACE THIS LINE WITH YOUR CODE CHUNK ###

```{r}
resume |> 
  group_by(firstname) |> 
  summarise(prop_callback = mean(call)) |> 
  arrange(-prop_callback)
```

