# Set up

For our first example, we will continue using data from Chetty et al's 2014 paper "Where Is The Land Of Opportunity?". The `commuting_zones.csv` file on Canvas comes from the Opportunity Insights website which can be accessed [`here`](http://www.opportunityinsights.org).

Load the data as a data frame called `cz` and load tidyverse and datatable.

```{r, warning = FALSE, error = FALSE, message = FALSE}
library(tidyverse)
library(DT)

cz <- read_csv("https://raw.githubusercontent.com/mjclawrence/soci385_f23/main/data/commuting_zones.csv")
```

# Introducing Multivariate Relationships

Last week we identified individual points on a scatterplot to dig deeper into how two variables are associated. Another analytical tool is to examine whether the association differs at specific values of another variable.

Consider the correlation between racial segregation and household income across commuting zones. Would you expect this correlation to be positive or negative? Strong or weak?

### REPLACE THIS LINE WITH YOUR CODE

Would this association be the same in urban and rural commuting zones?

### REPLACE THIS LINE WITH YOUR CODE


## Working With Lots Of Correlations

Let's move to a different example. Consider the relationship between Covid vaccination rates and eviction rates across zip codes. Would you expect a positive or negative association? How strong?

Load the data in a dataset called `evictions`:

```{r}
#| load eviction data
#| message: false

evictions <- read_csv("https://raw.githubusercontent.com/mjclawrence/soci385_f23/main/data/evictions.csv")
```


The dataset comes from the Eviction Lab and was provided by Olivia Jin, Middlebury alum and lead author of this paper:

Jin, Olivia, Emily Lemmerman, Peter Hepburn, and Matthew Desmond. 2021. "Neighborhoods with the Highest Eviction Filing Rates Have the Lowest Levels of COVID-19 Vaccination." *Socius* 7: 1-3.

What is the overall correlation between eviction rate (`evict_pct`) and vaccination rate (`vax_pct`)?

### REPLACE THIS LINE WITH YOUR CODE CHUNK ###

Now try using `group_by()` and `summarise()` to find how this association varies across the nine cities the researchers studied. The variable for cities is called `xsite`.

### REPLACE THIS LINE WITH YOUR CODE CHUNK ###

What are some different options for how to visualize these associations in each city?

The `facet_wrap()` option is a new way to create separate plots for different categories.

```{r}
#| label: facet_wrap example

evictions |> 
  ggplot(aes(x = evict_pct, y = vax_pct,
             color = xsite)) +
  geom_point() + 
  facet_wrap(~xsite) + # just the tilde (~), no period like facet_grid
  guides(color = "none") # drop the legend since each facet is titled
``` 

By default, `facet_wrap()` assigns each plot the same x-axis range and the same y-axis range. You can adjust them by adjusting the scales option with "free_x", "free_y" or "free" (for both the x and y axes):

```{r}
#| label: free scales with facet_wrap

evictions |> 
  ggplot(aes(x = evict_pct, y = vax_pct,
             color = xsite)) +
  geom_point() + facet_wrap(~xsite, scales = "free") +
  guides(color = "none")
```

How else to improve this figure?


## Adding Another Variable

This dataset also includes the "majority race" in each zip code. We know how to use `group_by()` and `summarise()` to get the correlation for each city. To get the correlation for each city *and* each majority race within that city, simply add both variables to the `group_by()` function:

```{r}
#| label: grouping by multiple variables

evictions |> 
  group_by(xsite, majority_race) |> # add variables with comma
  summarise(cor = cor(evict_pct, vax_pct,
                      use = "complete"))
```

This is a good time to use datatable...

```{r}
#| label: datatable after summarise

evictions |> 
  group_by(xsite, majority_race) |> # add variables with comma
  summarise(cor = cor(evict_pct, vax_pct,
                      use = "complete")) |> 
  datatable(rownames = FALSE)
```

Let's combine everything we know about plotting to add different colored points for each majority race. Start by changing the color aesthetic we used earlier...

```{r}
#| label: facet wrap with cities and majority races

evictions |> 
  ggplot(aes(x = evict_pct, y = vax_pct,
             color = majority_race)) +
  geom_point() + facet_wrap(~xsite, scales = "free") +
  guides(color = "none") + 
  geom_smooth(method = "lm")
```

How to improve this figure?

## Some Additional Thoughts About Color

One popular package to use for adjusting color is the R Color Brewer package. You can install it and load it here:

```{r}
#| label: install and load RColorBrewer

#install.packages("RColorBrewer") # hashtag this line after installing
library(RColorBrewer)
```

There's good information on Brewer's color options at [this site](http://colorbrewer2.org/#type=diverging&scheme=RdYlGn&n=3).

The site previews options for sequential (or continuous) variables and divergent (or factor/categorical/discrete) variables. Once you find a color scheme you like, input the palette name into the `scale_color_brewer()` function or the `scale_fill_brewer()` function.

```{r}
#| label: RColorBrewer example

evictions |> 
  ggplot(aes(x = evict_pct, y = vax_pct,
             color = xsite)) +
  geom_point() + facet_wrap(~xsite, scales = "free") +
  guides(color = "none") + 
  geom_smooth(method = "lm") +
  scale_color_brewer(palette = "Set1")
```

Want other colors? I like the [`viridis color palettes`](https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html) which we saw last week.

```{r}
#| label: Viridis example

evictions |> 
  ggplot(aes(x = evict_pct, y = vax_pct,
             color = xsite)) +
  geom_point() + facet_wrap(~xsite, scales = "free") +
  guides(color = "none") + geom_smooth(method = "lm") +
  scale_color_viridis_d(option = "mako") # Palette Name
```