---
title: "Assignment One"
author: "Matt Lawrence"
date: "September 26, 2023"
format: "html"
---

```{r, echo = FALSE, warning = FALSE, message = FALSE}
assignment1 <- read.csv("https://raw.githubusercontent.com/mjclawrence/soci385_f23/main/data/assignment_01.csv")
library(tidyverse)
```


### Questions

1. What are the mean and median of `agekdbrn`?

```{r}
summary(assignment1$agekdbrn)
```


2. Save the value of the 33rd percentile of `agekdbrn` as an object, save the value of the 67th percentile of `agekdbrn` as an object, and find the difference between these two values using the object names and R's built-in calculator function. In words, describe what this range tells us. ***[NOTE: Object names have to start with letters, not numbers!]***

```{r}
quantile(assignment1$agekdbrn, c(0.33, 0.67))
```

```{r}
q33 <- 21
q67 <- 27
q67 - q33
```

One-third of the sample has their first child in the 6 years between ages 21 and 27

3. What is the mode of `agekdbrn` for respondents who completed 12 or fewer years of education?

```{r}
sort(table(assignment1$agekdbrn[assignment1$educ<=12]), decreasing = TRUE)
```


4. What proportion of respondents completed exactly 16 years of education?

```{r}
prop.table(table(assignment1$educ))
```


5. Use tidyverse functions (remember to load the package first!) to create a new data frame with only the `agekdbrn` and `educ` variables, and that is limited to respondents who have 14 or more years of education. How many observations are in this new data frame?

```{r}
newdf <- assignment1 |> 
  select(agekdbrn, educ) |> 
  filter(educ>=14)
```


6. Use tidyverse functions to find the mean of `agekdbrn` for respondents with each year of highest schooling completed in this new data frame.

```{r}
newdf |> 
  group_by(educ) |> 
  summarise(mean = mean(agekdbrn))
```


7. Approximately how long did it take you to complete this assignment?
